---
title: "02_Recurrent_event_analysis"
format: html
editor: visual
---

## Modeling recurrent events in survival analysis

```{r}
library(survival)
```

The modeling of recurrent events is illustrated with the
bladder cancer dataset. Recurrent events are represented
in the data with multiple observations for subjects having
multiple events. The data layout for the bladder cancer
dataset is in the counting process (start, stop) format
with time intervals defined for each observation.

```{r}
# load data
bladder <- readRDS("../01_data/bladder.rds")
head(bladder, 20)
```

There are three observations for ID=10, one observation
for ID=11, three observations for ID=12, and two observations
for ID=13. The variables START and STOP represent
the time interval for the risk period specific to that observation. The (start,stop) data format provides a way to indicate
that a subject experienced multiple events.


Next, a recurrent-events Cox model is
run with the predictors: treatment status (TX), initial number
of tumors (NUM), and the initial size of tumors (SIZE).

```{r}
Y=Surv(bladder$start,bladder$stop,bladder$event==1)
coxph(Y ~ tx + num + size + cluster(id), data=bladder)
```

The estimated hazard ratio
(TX=1 vs. TX=0) is 0.663 (with a p-value of 0.0980) and hence not significant.

A stratified Cox model can also be run using the data in this
format with the variable INTERVAL as the stratified variable.
The stratified variable indicates whether the subject
was at risk for their first, second, third, or fourth event. This approach is called a Stratified CP recurrent event
model and is used if the investigator wants
to distinguish the order in which recurrent events occur.
The bladder data is in the proper format to run this model.

```{r}
coxph(Y ~ tx + num + size + strata(interval) + cluster
(id),data=bladder)
```
Another stratified approach (called Gap Time) is a slight
variation of the Stratified CP approach. The difference is in
the way the time intervals for the recurrent events are
defined. There is no difference in the time intervals when
subjects are at risk for their first event. However, with the
Gap Time approach, the starting time at risk gets reset to
zero for each subsequent event. To run a Gap Time model,
we need to create two new (start, stop) variables in the
bladder dataset, which we’ll call START2 and STOP2.

```{r}
bladder$start2=0
bladder$stop2=bladder$stop - bladder$start

Y2=Surv(bladder$start2,bladder$stop2,bladder$event)
coxph(Y2 ~ tx + num + size + strata(interval) + cluster(id),data=bladder)
```
The results using the Gap Time approach varies slightly
from that obtained using the Stratified CP approach.

